Geospatial - Assignment 3

### 2025-02-28 hw3/hw3_MoritzPeistNoemiLucchiSimonVellin.Rmd
exercise 2 part b code and documentation done

@@

  ![Map 1](data/ex2/raster07.png)

...

and the second graph as follows:

@@

>>>>

# Part A

@@

  ![Map 1](data/ex2/raster07.png)
</div>

# Part B - Logical Approach

Using the distance matrix calculated in Part A, we now assess the isolation of Madrid and Vigo by analyzing their bilateral distances to the other top 10 most populated cities in Spain. We begin by extracting the distances for Madrid and Vigo.

```{r}
# Extract city names from the top 10 cities (excluding Madrid and Vigo for isolation analysis)
other_cities <- spain_top_10_cities$NAME[!spain_top_10_cities$NAME %in% c("Madrid", "Vigo")]

# Extract distances from Madrid to other cities
madrid_distances <- dist_matrix["Madrid", other_cities]

# Extract distances from Vigo to other cities
vigo_distances <- dist_matrix["Vigo", other_cities]
```

## Analysis

Isolation is determined by comparing the average distances from each city to the others, where a higher average distance indicates greater isolation. We compute summary statistics, and calculate the average distance for each city to make a quantitative comparison.

```{r}
# Calculate summary statistics for isolation
madrid_summary <- summary(madrid_distances)
vigo_summary <- summary(vigo_distances)

# Print summary statistics to compare isolation
cat("\nSummary of distances from Madrid to other top 10 cities (km):\n")
print(round(madrid_summary, 1))
cat("\nSummary of distances from Vigo to other top 10 cities (km):\n")
print(round(vigo_summary, 1))

# Calculate average distances to assess isolation
madrid_avg_distance <- mean(madrid_distances)
vigo_avg_distance <- mean(vigo_distances)

cat("\nAverage distance from Madrid to other cities:", round(madrid_avg_distance, 1), "km\n")
cat("Average distance from Vigo to other cities:", round(vigo_avg_distance, 1), "km\n")

# Determine which city is more isolated (higher average distance indicates more isolation)
if (madrid_avg_distance > vigo_avg_distance) {
  cat("Madrid appears more isolated based on average distance.\n")
} else if (vigo_avg_distance > madrid_avg_distance) {
  cat("Vigo appears more isolated based on average distance.\n")
} else {
  cat("Madrid and Vigo have similar isolation based on average distance.\n")
}
```

To further explore the isolation of Madrid and Vigo, we create smoothed density distributions of their distances to the other top 10 cities. This visualization helps us understand the spread and concentration of distances, where a distribution with a wider spread or higher density at greater distances suggests greater isolation. We use geom_density() to plot the distributions, with Madrid in red and Vigo in turquoise, matching the style of the reference plot.

```{r}
# Create a data frame with distances for plotting
density_data <- data.frame(
  City = c(rep("Madrid", length(madrid_distances)), rep("Vigo", length(vigo_distances))),
  Distance = c(madrid_distances, vigo_distances)
)

# Plot smoothed density distributions using geom_density()
density_plot <- ggplot(density_data, aes(x = Distance, color = City, fill = City)) +
  geom_density(alpha = 0.3) +  # Add transparency for overlap
  scale_color_manual(values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  scale_fill_manual(values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  labs(
    title = "Smoothed Density Distributions of Distances",
    x = "Distance (km)",
    y = "Density",
    color = "City",
    fill = "City"
  ) +
  theme_minimal()

# Display the plot
print(density_plot)
```

In comparison the second graph looked as follows:

@@

# Part B - Alternative Approach

Although we think it seems logical to exclude Madrid and Virgo from the distance calculations, the reference plot seem to include them, as displayed by the noticeable density at x=0 for both cities. In this section, we computed the same plot but with "self-distances" included.

```{r}
# Extract distances for Madrid and Vigo to all cities, including themselves (self-distance = 0)
madrid_distances_all <- dist_matrix["Madrid", ]
vigo_distances_all <- dist_matrix["Vigo", ]

# Create a data frame with distances for plotting, including self-distances
density_data_all <- data.frame(
  City = c(rep("Madrid", length(madrid_distances_all)), rep("Vigo", length(vigo_distances_all))),
  Distance = c(madrid_distances_all, vigo_distances_all)
)

# Plot smoothed density distributions using geom_density(), including self-distances
density_plot_all <- ggplot(density_data_all, aes(x = Distance, color = City, fill = City)) +
  geom_density(alpha = 0.3) +  # Add transparency for overlap
  scale_color_manual(values = c("Madrid" = "red", "Vigo" = "turquoise"), labels = c("Madrid", "Vigo")) +
  scale_fill_manual(values = c("Madrid" = "red", "Vigo" = "turquoise"), labels = c("Madrid", "Vigo")) +
  labs(
    x = "distance",
    y = "density",
    color = "origin",
    fill = "origin"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

# Display the plot
print(density_plot_all)
```
